---
title: "DATA 304 Gallery"
author: "Lily McAboy"
format:
  html:
    embed-resources: true
    code-tools: true
---
```{r}
#| include: false
library(vegawidget)
library(dplyr)
library(reticulate)
library(altair)
```

```{python}
#| include: false
import pandas as pd
import vega_datasets as data
import altair as alt
```

# Exercise 1

Copy of Graphic: 

![DataIsBeautiful](https://www.reddit.com/r/dataisbeautiful/comments/1ij5q7c/oc_three_worst_singleday_stock_market_drops_in)

What marks are being used? What variables are mapped to which properties?

What is the main story of this graphic?

What makes it a good graphic?

What features do you think you would know how to implement in Vega-Lite?

Are there any features of the graphic that you would not know how to do in Vega-Lite? If so, list them.

# Exercise 2

## Visualizing Weather 1
Create a graphic that shows the mean temperature for each month. How many “months” should you be displaying? (There is more than one answer to this – perhaps try doing it more that one way.)

### Way 1

In this graphic, I am showing the mean max temperature per month, color is used as a guide to show the year for a specific month. 

```{r}
'
{
"$schema": "https://vega.github.io/schema/vega-lite/v5.json",
"data": {
  "url": "https://calvin-data304.netlify.app/data/weather-with-dates.csv"
},
"height": 400,
"width": 500,
"title": "Mean Max Temperature Per Month",
"mark": "circle" ,
"encoding": {
  "x": {"field": "month", "type": "temporal", "timeUnit": "month","title": "Month"},
  "y": {"field": "temp_max", "type": "quantitative", "aggregate": "average", "title": "Temperature (F)"},
  "color": {"field": "year", "title": "Year"},
  "fillOpacity": {"value": 0.8},
  "size": {"value": 80}
}
}' |> as_vegaspec()
```
### Way 2

Using Python and Pandas to calculate the actual mean temperature for each day, and then aggregating by month

```{python}

```

```{python}

```



## Visualizing Weather 2
Exercise 3 Create a graphic that shows how the different types of weather (rain, fog, etc.) are distributed by month in Seattle. When is it rainiest in Seattle? Sunniest?

